--- 
title: "Back to Basics: Refresh Your R Skills" 
format: html 
toc: true 
toc-location: right 
--- 

Don’t we all know it? Summer ends and we’ve forgotten where we left off before the break. Has this happened with your R programming? No problem! In this session, we’ll refresh your R skills and give you a boost for the upcoming academic year. On 6th of October, we will cover the technical basics (basic syntax, how to run R scripts, and how to install and load packages), simple data processing (how to import and explore your data), and basic visualization (using base plots and ggplot2).

# The technical basics

## Basic syntax

### Naming objects

You can use R similar to a calculator. It is possible to just type in numbers into your console and receive results immediatly.

```{r}
9 * 5
```

Everything that you create (and that exists) in R is an object. To be able to use them efficiently, we can name objects. For that, we use the assignment operator `<-`.

```{r}
result <- 9 * 5
```

If we want to print the value of an object, you can run a line that just contains the object name.

```{r}
result
```

### Comments
Comments are a programmers friend. They help you to keep an overview of your code, understand what you did (and why) and keep your code readable. In R, you can use the `#` character to create a comment. You can do this either behind a line of code or above or below it.

```{r}
# This is a comment
code <- 10 * 3 # I can also add a comment here
# This is also a comment
```

You can also use comments to exclude specific lines of code from being run. To do that, you "comment out" a line of code. In the following example, we commented the line out that assigns `b` the new value `20`.

```{r}
a <- 30
b <- 100
# b <- 20 

c <- a * b
c
```

### Functions and arguments
R offers a various amounts of in-built functions, i.e., a block of code that runs when the function is used. To use a function, we use its name and provide it with the needed arguments, i.e., the input that the function needs to run.
In the following example, we use the function `sqrt` to calculate the square root of 4.
```{r}
sqrt(4)
```

If you want to know more about a function, you can use the `?` to access the documentation.
```{r}
?sqrt()
```

### Vectors and data types
In R, vectors are a commonly used data type. Vectors consists of a series of values which can be connected using the `c()`-function.

```{r}
my_vector <- c(1,2,3,4,5)
my_vector
```

I can also store characters in a vector. Characters are indicated with `""`.
```{r}
my_second_vector <- c("these","are","characters")
my_second_vector
```

Aside from `numeric` and `character` data types, R also allows to use `logical` data (`TRUE` vs. `FALSE`), `integer`, `complex` and `raw`.  

You can use various function on vectors (and even subset them). Commonly used are the following:
```{r}
length(my_vector) # returns the length of the vector
typeof(my_vector) # returns the type of the vector (e.g., character)
str(my_vector) # returns an overview of the structure of the vector
```

## How to run R scripts
You can run R code either through the command line or by using an R script. Usually, you run the code line by line. An example of an R script can be found in [our Programming Café session on IDEs](https://eur-nl.github.io/rs_training/IDEs.html#get-to-know-rstudio) in which we present how RStudio can be used to run R scripts.

## How to install and load packages

To install new packages, you can use the command `install.packages("PACKAGE")`. To import packages to use them in your program, use `library(PACKAGE)`. The following example shows how to install and import ggplot2.
```{{r}}
install.packages("ggplot2")
library(ggplot2)
```

# Simple data analyses

## How to import your data
You can import your data with the help of various packages. The most straight-forward way is to use one of the following functions:
```{r}
# Read file in table format (each row is one line in the document)
data <- read.table(file = "data/documents.csv", header = TRUE, sep = "\t")

# import file with sep = "\t"
data <- read.delim(file = "data/documents.csv") 
# import .csv file
data <- read.csv(file = "data/documents.csv") 

```

## How to explore your data

You can have a look at your data with various in-built functions. The following overview is taken from the [Data Carpentry for SSH workshop](https://datacarpentry.github.io/r-socialsci/02-starting-with-data.html#importing-data) and adapted to fit our example:

### Size
```{r}
dim(data) # - returns a vector with the number of rows as the first element, and the number of columns as the second element (the dimensions of the object)
nrow(data) # - returns the number of rows
ncol(data) # - returns the number of columns
```

### Content
```{r}
head(data) #- shows the first 6 rows
tail(data) #- shows the last 6 rows
```

### Names
```{r}
names(data) #- returns the column names (synonym of colnames() for data.frame objects)
```

### Summary
```{r}
str(data) #- structure of the object and information about the class, length and content of each column
summary(data) #- summary statistics for each column
```

## How to use characters as factors
As most of our data consists of characters, we might want to make some of them categorical variables. We can do that by casting them as factors using the `as.factor()`-function. We address the specific column using `$`.

```{r}
data$Gender <- as.factor(data$Gender)
data$Complaint <- as.factor(data$Complaint)
data$ZIP.Code <- as.factor(data$ZIP.Code)
```

## How to create a base plot
Creating a base plot in R is as easy as typing `plot(data)`. 
```{r}
plot(data)
```

For our data, the plot is not really helpful, but you can also focus on one specfic column instead using `$` to indicate the column name. 

```{r}
plot(data$Gender)
```

`plot()` allows for multiple other arguments to be used. You can find more information on it [here](https://eur-nl.github.io/rs_training/DataVisualization.html).

A handy package to create plots is `ggplot2`. You can find a manual on how to visualize your data with it [here](https://eur-nl.github.io/rs_training/DataVisualization.html).


# Further reading